---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sosprosody

<!-- badges: start -->
  [![Codecov test coverage](https://codecov.io/gh/tsostarics/sosprosody/branch/main/graph/badge.svg)](https://app.codecov.io/gh/tsostarics/sosprosody?branch=main)
  <!-- badges: end -->

This package includes a variety of helper functions involving praat, textgrids,
pitchtiers, and intonation-related data wrangling.
This has been written for my own research, but others may find it helpful as
well.
This package extends the `rPraat` package.

## Installation

You can install the development version of `sosprosody` like so:

``` r
library(devtools)
devtools::install_github('tsostarics/sosprosody')
```

## Examples

This package provides `format` and `print` methods for TextGrid and PitchTier
objects loaded with the respective read functions from `{rPraat}`

```{r}
#| warning: false
library(rPraat)
library(sosprosody)
pitchtier <- pt.read("grandmother_LHH_003.PitchTier")
textgrid <- tg.read("grandmother_LHH_003.TextGrid")

pitchtier
textgrid
```

Here's an example of processing all the textgrids and pitchtiers in a directory
into dataframe representations.

```{r}
nuclear_words <- "grandmother"
tg_df <- batch_process_textgrids("./")
pt_df <- batch_process_pitchtiers("./")
nuclear_regions <- get_nuclear_textgrids(tg_df, nuclear_words)

str(tg_df)
str(pt_df)
```

We can then do some common preprocessing steps, such as coding the nuclear word
in the phrase, applying running median smoothing, and normalizing the timepoints.

```{r}
processed_pt_df <- preprocess_pitchtracks(pt_df,
                                          nuclear_df = nuclear_regions, 
                                          runmed_k = 5,
                                          time_normalize = TRUE,
                                          .fromzero = TRUE)
```

The processed pitch tier data frame can then be plotted like so:

```{r}
library(ggplot2)

processed_pt_df |> 
  ggplot(aes(x = timepoint_norm, 
             y = hz_runmed, 
             color = is_nuclear,
             group = file)) +
  geom_point() +
  geom_line() +
  # The rest is just for visuals
  scale_color_brewer(palette = 'Dark2') +
  theme_minimal() +
  coord_fixed(ratio = 1/180) +
  theme(legend.position = 'none') +
  annotate(geom = "label", x = .9, y = 85, label = "L*+HLL") +
  annotate(geom = "label", x = .75, y = 150, label = "LHH")
```   
 
 
